---
title: "Quarto Test Playground"
author: "Your Name"
date: "2026-02-17"
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
---

## Overview

This file includes multiple code cells so you can test rendering behavior.

```{r}
#| label: setup
#| include: false
set.seed(123)
```

## Cell 1: Basic Math

```{r}
#| label: basic-math
x <- 1:10
mean(x)
sd(x)
```

## Cell 2: Objects and Printing

```{r}
#| label: object-print
name <- "Quarto"
version_test <- TRUE
list(name = name, working = version_test)
```

## Cell 3: Data Frame

```{r}
#| label: create-data
sample_df <- data.frame(
  id = 1:6,
  group = c("A", "A", "B", "B", "C", "C"),
  score = c(87, 92, 75, 81, 95, 88)
)
sample_df
```

## Cell 4: dplyr Example (optional)

```{r}
#| label: dplyr-example
# If dplyr is installed, this will run.
if (requireNamespace("dplyr", quietly = TRUE)) {
  dplyr::as_tibble(sample_df) |>
    dplyr::group_by(group) |>
    dplyr::summarise(avg_score = mean(score), .groups = "drop")
} else {
  "Install dplyr to run this chunk: install.packages('dplyr')"
}
```

## Cell 5: Plot

```{r}
#| label: scatter-plot
#| fig-width: 6
#| fig-height: 4
plot(cars$speed, cars$dist,
     pch = 19,
     col = "steelblue",
     main = "cars dataset",
     xlab = "Speed",
     ylab = "Distance")
abline(lm(dist ~ speed, data = cars), col = "firebrick", lwd = 2)
```

## Cell 6: Histogram

```{r}
#| label: histogram
hist(rnorm(200),
     breaks = 20,
     col = "gray80",
     border = "white",
     main = "Random Normal Values")
```

## Cell 7: Chunk Option Demo (no echo)

```{r}
#| label: hidden-code-output
#| echo: false
paste("Current date from R:", Sys.Date())
```

## Cell 8: Intentional Warning Demo

```{r}
#| label: warning-demo
#| warning: true
log(-1)
```

## Cell 9: Intentional Error Demo (set error=false to test failure)

```{r}
#| label: error-demo
#| error: true
stop("This is a test error chunk.")
```

## Markdown Table

| Item | Value |
|---|---:|
| A | 10 |
| B | 20 |
| C | 30 |

## Try Rendering

Run this in terminal:

```bash
quarto render ~/Desktop/test.qmd
```

Or render from R:

```{r}
#| eval: false
quarto::quarto_render("~/Desktop/test.qmd")
```
