---
title: "Assignment #1"
format: html
---

## Assignment 1

You only need to write lines of code for each question. When answering questions that ask you to identify or interpret something, the length of your response doesn’t matter. For example, if the answer is just ‘yes,’ ‘no,’ or a number, you can just give that answer without adding anything else.

We will go through comparable code and concepts in the live learning session. If you run into trouble, start by using the help help() function in R, to get information about the datasets and function in question. The internet is also a great resource when coding (though note that no outside searches are required by the assignment!). If you do incorporate code from the internet, please cite the source within your code (providing a URL is sufficient).

Please bring questions that you cannot work out on your own to office hours, work periods or share with your peers on Slack. We will work with you through the issue.

You will need to install PLINK and run the analyses. Please follow the OS-specific setup guide in [`SETUP.md`](../../SETUP.md). PLINK is a free, open-source whole genome association analysis toolset, designed to perform a range of basic, large-scale analyses in a computationally efficient manner.

#### Question 1: Data inspection

Before fitting any models, it is essential to understand the data. Use R or bash code to answer the following questions about the `gwa.qc.A1.fam`, `gwa.qc.A1.bim`, and `gwa.qc.A1.bed` files, available at the following Google Drive link: <https://drive.google.com/drive/folders/11meVqGCY5yAyI1fh-fAlMEXQt0VmRGuz?usp=drive_link>. Please download all three files from this link and place them in `02_activities/data/`.
```{r}

fam_file <- "../data/gwa.qc.A1.fam"
bim_file <- "../data/gwa.qc.A1.bim"
bed_file <- "../data/gwa.qc.A1.bed"


fam_data <- read.table(fam_file, header = FALSE)
cat("Number of samples:", nrow(fam_data), "\n")
head(fam_data)

bim_data <- read.table(bim_file, header = FALSE)
cat("Number of SNPs:", nrow(bim_data), "\n")
head(bim_data)

library(snpStats)

plink_data <- read.plink(
  bed = bed_file,
  bim = bim_file,   # pass path, NOT the data frame
  fam = fam_file    # pass path, NOT the data frame
)

geno_matrix <- plink_data$genotypes
cat("Genotype matrix dimensions (samples x SNPs):", dim(geno_matrix), "\n")
```

(i) Read the .fam file. How many samples does the dataset contain?


```{r}
ffam_file <- "../data/gwa.qc.A1.fam"
bim_file <- "../data/gwa.qc.A1.bim"
bed_file <- "../data/gwa.qc.A1.bed"
```
```{r}
# Read the data file
data <- read.table("../data/gwa.qc.A1.fam", header = FALSE)

# Display number of rows
nrow(data)

# Display first few rows
head(data)
```







```

(ii) What is the 'variable type' of the response variable (i.e.Continuous or binary)?

```
```{r}
data <- read.table(fam_file, header = FALSE)
```
```{r}
table(data$V6)
```

(iii) Read the .bim file. How many SNPs does the dataset contain?


```{r}       
# Read the .bim file
bbim_data <- read.table(bim_file, header = FALSE)
# Number of SNPs
num_snps <- nrow(bim_data)
cat("The dataset contains", num_snps, "SNPs.\n")
```

#### Question 2: Allele Frequency Estimation

(i) Load the genotype matrix for SNPs rs1861, rs3813199, rs3128342, and rs11804831 using additive coding. What are the allele frequencies (AFs) for these four SNPs?
```{r}
 # Load required library
library(snpStats)

# Read PLINK files
plink_data <- read.plink(
  bed = bed_file,
  bim = bim_file,
  fam = fam_file
)

# Extract genotype matrix
geno_matrix <- plink_data$genotypes

# Subset SNPs of interest
snps_of_interest <- c("rs1861", "rs3813199", "rs3128342", "rs11804831")
geno_sub <- geno_matrix[, snps_of_interest]

# Convert to numeric (0,1,2)
geno_numeric <- as(geno_sub, "numeric")

# Function to calculate allele frequency
calc_af <- function(snp_vector) {
  sum(snp_vector, na.rm = TRUE) / (2 * sum(!is.na(snp_vector)))
}

# Compute allele frequencies
allele_freqs <- apply(geno_numeric, 2, calc_af)

# Print results
allele_freqs
```

(ii) What are the minor allele frequencies (MAFs) for these four SNPs?

``` {r}        
# Function to calculate minor allele frequency
calc_maf <- function(snp_vector) {
  af <- sum(snp_vector, na.rm = TRUE) / (2 * sum(!is.na(snp_vector)))  # allele freq
  maf <- min(af, 1 - af)  # minor allele freq
  return(maf)
}

# Compute MAFs for all SNPs in geno_numeric
maf_values <- apply(geno_numeric, 2, calc_maf)

# Print results
maf_values
```

#### Question 3: Hardy–Weinberg Equilibrium (HWE) Test

(i) Conduct the Hardy–Weinberg Equilibrium (HWE) test for all SNPs in the .bim file. Then, load the file containing the HWE p-value results and display the first few rows of the resulting data frame.

```  {r}       
hwe_results <- read.table("../data/gwa_hwe.hwe", header = TRUE)
head(hwe_results)
```

(ii) What are the HWE p-values for SNPs rs1861, rs3813199, rs3128342, and rs11804831?

``` {r}      

library(dplyr)

# Now filter SNPs
snps_of_interest <- c("rs1861", "rs3813199", "rs3128342", "rs11804831")

hwe_results %>%
  filter(SNP %in% snps_of_interest) %>%
  select(SNP, P)
```

#### Question 4: Genetic Association Test

(i) Conduct a linear regression to test the association between SNP rs1861 and the phenotype. What is the p-value?


```{r}
# Read in the phenotype from .fam
fam_data <- read.table("../data/gwa.qc.A1.fam", header = FALSE)
geno_matrix <- as(geno_matrix, "numeric")  # make sure numeric

# Add phenotype to genotype subset
geno_df <- as.data.frame(geno_numeric)
geno_df$PHENOTYPE <- fam_data$V6  # column 6 is phenotype

# Fit linear regression for rs1861
model <- lm(PHENOTYPE ~ rs1861, data = geno_df)

# Summary to get p-value
summary(model)$coefficients

# Extract p-value for rs1861
p_value <- summary(model)$coefficients["rs1861", "Pr(>|t|)"]

# Print it
cat("SNP: rs1861\nLinear regression p-value:", p_value, "\n")
```

(ii) How would you interpret the beta coefficient from this regression?

```{r}
summary(model)$coefficients
# Extract beta (estimate) and p-value for rs1861
beta <- summary(model)$coefficients["rs1861", "Estimate"]
pval <- summary(model)$coefficients["rs1861", "Pr(>|t|)"]

cat("rs1861: Beta =", beta, ", p-value =", pval, "\n")

cat ("The beta coefficient for rs1861 is approximately 0.974, indicating that each additional copy of the minor allele is associated with an increase of about 0.974 units in the phenotype under an additive model.\n")
```
(iii) Plot the scatterplot of phenotype versus the genotype of SNP rs1861. Add the regression line to the plot.

```{r} 
library(ggplot2)
library(dplyr)

# Create prediction data (smooth curve)
geno_range <- data.frame(
  rs1861 = seq(0, 2, length.out = 100)
)

geno_range$predicted <- predict(model, newdata = geno_range)

# Prepare count data (for point sizes)
count_data <- geno_df %>%
  mutate(PHENOTYPE = round(PHENOTYPE)) %>%
  filter(!is.na(rs1861), !is.na(PHENOTYPE)) %>%
  group_by(rs1861, PHENOTYPE) %>%
  summarise(count = n(), .groups = "drop")

# Plot
p <- ggplot() +
  geom_point(
    data = count_data,
    aes(x = rs1861, y = PHENOTYPE, size = count),
    color = "blue"
  ) +
  geom_line(
    data = geno_range,
    aes(x = rs1861, y = predicted),
    color = "red",
    size = 1.2
  ) +
  scale_size_continuous(range = c(2, 10)) +
  labs(
    title = "Phenotype vs Genotype (rs1861)",
    x = "Genotype (0/1/2)",
    y = "Phenotype",
    size = "Count"
  ) +
  coord_cartesian(xlim = c(-0.5, 2.5), ylim = c(-0.1, 1.1)) +
  theme_minimal()

print(p)
```

(iv) Convert the genotype coding for rs1861 to recessive coding.

``` {r}      
# Create a new variable with recessive coding
geno_df$rs1861_rec <- ifelse(geno_df$rs1861 == 2, 1, 0)

# Check result
table(geno_df$rs1861, geno_df$rs1861_rec)
```

(v) Conduct a linear regression to test the association between the recessive-coded rs1861 and the phenotype. What is the p-value?

```{r}
# Fit linear regression with recessive coding
model_rec <- lm(PHENOTYPE ~ rs1861_rec, data = geno_df)

# Extract p-value
p_value_rec <- summary(model_rec)$coefficients["rs1861_rec", "Pr(>|t|)"]

# Print result
cat("Recessive model p-value:", p_value_rec, "\n")
```

(vi) Plot the scatterplot of phenotype versus the recessive-coded genotype of rs1861. Add the regression line to the plot.

``` {r}       
# Scatterplot
plot(geno_df$rs1861_rec, geno_df$PHENOTYPE,
     xlab = "Recessive Genotype (rs1861)",
     ylab = "Phenotype",
     main = "Phenotype vs Recessive rs1861",
     pch = 16)

# Add regression line
abline(model_rec, col = "red", lwd = 2)
```

(vii) Which model fits better? Justify your answer.

```{r}       
# R-squared for additive model
summary(model)$r.squared

# R-squared for recessive model
summary(model_rec)$r.squared

# Compare model fits
cat("The additive model fits slightly better than the recessive model because its R-squared is higher (0.0892 vs 0.0858), even though both p-values are highly significant.\n")
```


### Criteria

| Criteria | Complete | Incomplete |
|------------------------|------------------------|------------------------|
| **Data Inspection** | Correct sample/SNP counts and variable type identified. | Missing or incorrect counts or variable type. |
| **Allele Frequency Estimation** | Correct allele and minor allele frequencies computed. | Frequencies missing or wrong. |
| **Hardy–Weinberg Equilibrium Test** | Correct PLINK command and p-value extraction in R. | PLINK command or extraction incorrect/missing. |
| **Genetic Association Test** | Correct regressions, plots, coding, and interpretation. | Regression, plots, or interpretation missing/incomplete. |

### Submission Information

📌 Please review our [Assignment Submission Guide](https://github.com/UofT-DSI/onboarding/blob/main/onboarding_documents/submissions.md) for detailed instructions on how to format, branch, and submit your work. Following these guidelines is crucial for your submissions to be evaluated correctly.

#### Note:

If you like, you may collaborate with others in the cohort. If you choose to do so, please indicate with whom you have worked with in your pull request by tagging their GitHub username. Separate submissions are required.

------------------------------------------------------------------------

#### Submission Parameters

-   Submission Due Date: `11:59 PM – 16/03/2026`

-   Branch name for your repo should be: `assignment-1`

-   What to submit for this assignment:

    -   Populate this Quarto document (`assignment_1.qmd`).
    -   Render the document with Quarto: `quarto render assignment_1.qmd`.
    -   Submit both `assignment_1.qmd` and the rendered HTML file `assignment_1.html` in your pull request.

-   What the pull request link should look like for this assignment: `https://github.com/<your_github_username>/gen_data/pull/<pr_id>`

    -   Open a private window in your browser. Copy and paste the link to your pull request into the address bar. Make sure you can see your pull request properly. This helps the technical facilitator and learning support team review your submission easily.

------------------------------------------------------------------------

Checklist:

-   Created a branch with the correct naming convention.
-   Ensured that the repository is public.
-   Reviewed the PR description guidelines and adhered to them.
-   Verified that the link is accessible in a private browser window.
-   Confirmed that both `assignment_1.qmd` and `assignment_1.html` are included in the pull request.

If you encounter any difficulties or have questions, please don't hesitate to reach out to our team via our Slack help channel. Our technical facilitators and learning support team are here to help you navigate any challenges.
