---
title: "Assignment #1"
format: html
---

## Assignment 1

You only need to write lines of code for each question. When answering questions that ask you to identify or interpret something, the length of your response doesn’t matter. For example, if the answer is just ‘yes,’ ‘no,’ or a number, you can just give that answer without adding anything else.

We will go through comparable code and concepts in the live learning session. If you run into trouble, start by using the help help() function in R, to get information about the datasets and function in question. The internet is also a great resource when coding (though note that no outside searches are required by the assignment!). If you do incorporate code from the internet, please cite the source within your code (providing a URL is sufficient).

Please bring questions that you cannot work out on your own to office hours, work periods or share with your peers on Slack. We will work with you through the issue.

You will need to install PLINK and run the analyses. Please follow the OS-specific setup guide in [`SETUP.md`](../../SETUP.md). PLINK is a free, open-source whole genome association analysis toolset, designed to perform a range of basic, large-scale analyses in a computationally efficient manner.

#### Question 1: Data inspection

Before fitting any models, it is essential to understand the data. Use R or bash code to answer the following questions about the `gwa.qc.A1.fam`, `gwa.qc.A1.bim`, and `gwa.qc.A1.bed` files, available at the following Google Drive link: <https://drive.google.com/drive/folders/11meVqGCY5yAyI1fh-fAlMEXQt0VmRGuz?usp=drive_link>. Please download all three files from this link and place them in `02_activities/data/`.

```{r, message=FALSE, warning=FALSE, results='hide'}
# Load the packages needed for this assignment
library(data.table)
library(ggplot2)
library(seqminer)
library(HardyWeinberg)
library(dplyr)
```

(i) Read the .fam file. How many samples does the dataset contain?
     
```{bash}
wc -l ../data/gwa.qc.A1.fam
```

The fam dataset contains 4000 samples

(ii) What is the 'variable type' of the response variable (i.e.Continuous or binary)?

```{bash}
head ../data/gwa.qc.A1.fam
```

The variable type or phenotype is continuous

(iii) Read the .bim file. How many SNPs does the dataset contain?

```{bash}
wc -l ../data/gwa.qc.A1.bim         
```

The bim file has 101083 SNPs

#### Question 2: Allele Frequency Estimation

(i) Load the genotype matrix for SNPs rs1861, rs3813199, rs3128342, and rs11804831 using additive coding. What are the allele frequencies (AFs) for these four SNPs?

```{bash,results='hide',message=FALSE, warning=FALSE}

# Create SNP list
printf "%s\n" rs1861 rs3813199 rs3128342 rs11804831 > ../data/snplist_A1.txt
cat ../data/snplist_A1.txt

# Subset the 4 SNPs from the PLINK dataset
plink2 --bfile ../data/gwa.qc.A1 --extract ../data/snplist_A1.txt --make-bed --out ../data/gwa_A1_subset

# Additive coding on the subsetted SNPs
plink2 --bfile ../data/gwa_A1_subset --export A --out ../data/gwa_A1_subset_additive

# Calculate allele frequencies for the 4-SNP subset
plink2 --bfile ../data/gwa_A1_subset --freq --out ../data/gwa_A1_subset_freq
```

```{r,message=FALSE, warning=FALSE}
# Load additive-coded genotype matrix
geno <- fread("../data/gwa_A1_subset_additive.raw")
head(geno)
```

```{r,message=FALSE, warning=FALSE}
# Read and display allele frequencies of four SNPs
freq <- fread("../data/gwa_A1_subset_freq.afreq")
af_table <- freq[, .(SNP = ID, AF = ALT_FREQS)]
af_table
```

The allele frequencies (AF) of the four SNPs are as follows: rs1861 = 0.0539859, rs3813199 = 0.0569126, rs3128342 = 0.3051210, rs11804831 = 0.1543410

(ii) What are the minor allele frequencies (MAFs) for these four SNPs?

```{r,message=FALSE, warning=FALSE}
maf_table <- freq[, .(SNP = ID, AF = ALT_FREQS, MAF = pmin(ALT_FREQS, 1 - ALT_FREQS))]
maf_table
```

Since the estimated allele frequencies of these SNPs are <0.5 from the ALT_FREQS column of the PLINK frequency output, ALT is already the minor allele. Therefore, ALT_FREQS = minor allele frequencies.

#### Question 3: Hardy–Weinberg Equilibrium (HWE) Test

(i) Conduct the Hardy–Weinberg Equilibrium (HWE) test for all SNPs in the .bim file. Then, load the file containing the HWE p-value results and display the first few rows of the resulting data frame.

```{bash,results='hide',message=FALSE, warning=FALSE}
plink2 --bfile ../data/gwa.qc.A1 --hardy --out ../data/gwa_qc_A1_hwe
```

```{r, message=FALSE, warning=FALSE}
hwe <- fread("../data/gwa_qc_A1_hwe.hardy")
head(hwe)  
```

(ii) What are the HWE p-values for SNPs rs1861, rs3813199, rs3128342, and rs11804831?

```{r, message=FALSE, warning=FALSE}     
# Create a subset of the four SNPs
snps_interest <- c("rs1861", "rs3813199", "rs3128342", "rs11804831")
hwe_subset <- hwe[ID %in% snps_interest, .(SNP = ID, HWE_P = P)]
hwe_subset
```

The HWE p-values for the four SNPs are as follows: rs1861 = 0.274719, rs3813199 = 1.000000, rs3128342 = 0.330273, rs11804831 = 0.113354

#### Question 4: Genetic Association Test

(i) Conduct a linear regression to test the association between SNP rs1861 and the phenotype. What is the p-value?

```{r, message=FALSE, warning=FALSE}         
# Linear regression model
geno <- fread("../data/gwa_A1_subset_additive.raw")
model <- lm(PHENOTYPE ~ rs1861_C, data = geno)
summary(model)
```

The p-value for the association between SNP rs1861 and the phenotype is <2e-16.

(ii) How would you interpret the beta coefficient from this regression?

The regression coefficient for rs1861 is 0.97382 (p <2e-16), indicating that each additional copy of the C allele is associated with an increase of approximately 0.97 units of the phenotype on average. The significant p-value suggests evidence of an association between this SNP and the phenotype in our sample.

(iii) Plot the scatterplot of phenotype versus the genotype of SNP rs1861. Add the regression line to the plot.

```{r, message=FALSE, warning=FALSE}          
# Load the genotype data
geno <- fread("../data/gwa_A1_subset_additive.raw")

# Remove missing values before plotting
plot_data_add <- geno[!is.na(rs1861_C) & !is.na(PHENOTYPE), ]

# Create scatterplot with regression line assuming outcome is phenotype and predictor is genotype of SNP rs1861
p <- ggplot(plot_data_add, aes(x = factor(rs1861_C), y = PHENOTYPE)) +
  geom_point(color = "blue", alpha = 0.5) +
  geom_smooth(aes(group = 1), method = "lm", color = "red") +
  labs(
    title = "Scatterplot showing the association between SNP rs1861 and phenotype",
    x = "Genotype (number of C alleles)",
    y = "Phenotype"
  ) +
  theme_minimal()

p
```

(iv) Convert the genotype coding for rs1861 to recessive coding.

```{r, message=FALSE, warning=FALSE}            
# Load genotype data
geno <- fread("../data/gwa_A1_subset_additive.raw")

# Convert coding for rs1861 to recessive coding
geno_recessive <- geno
geno_recessive$rs1861_recessive <- ifelse(geno_recessive$rs1861_C == 2, 1, 0)

# Check recoding
table(geno_recessive$rs1861_C, geno_recessive$rs1861_recessive, useNA = "ifany")
```

(v) Conduct a linear regression to test the association between the recessive-coded rs1861 and the phenotype. What is the p-value?

```{r, message=FALSE, warning=FALSE}         
# Linear regression model using recessive coding
model_recessive <- lm(PHENOTYPE ~ rs1861_recessive, data = geno_recessive)
summary(model_recessive)
```

The p-value for the association between SNP rs1861 and the phenotype using recessive coding is <2e-16.

(vi) Plot the scatterplot of phenotype versus the recessive-coded genotype of rs1861. Add the regression line to the plot.

```{r, message=FALSE, warning=FALSE}         
# Create scatterplot with regression line assuming outcome is phenotype and predictor is genotype of SNP rs1861 using recessive coding
plot_data_rec <- geno_recessive[!is.na(rs1861_recessive) & !is.na(PHENOTYPE), ]

p_rec <- ggplot(plot_data_rec,
                aes(x = factor(rs1861_recessive), y = PHENOTYPE)) +
  geom_point(color = "blue", alpha = 0.5) +
  geom_smooth(aes(group = 1), method = "lm", color = "red") +
  labs(
    title = "Scatterplot showing the association between recessive-coded rs1861 \nand phenotype",
    x = "Recessive genotype (1 = two copies of allele)",
    y = "Phenotype"
  ) +
  theme_minimal()

p_rec
```

(vii) Which model fits better? Justify your answer.

```{r, message=FALSE, warning=FALSE}
AIC(model)
AIC(model_recessive)
```

The additive model fits better than the recessive model. In the linear regression analyses, the additive model had a higher R-squared value compared to the recessive model (0.08924 vs. 0.08582). Furthermore, the AIC of the additive model (11276.91) is lower than the AIC of the recessive model (11291.74), indicating that the additive coding provides a better overall model fit. This suggests that the effect of rs1861 on the phenotype is better captured by an allele-dose (additive) effect rather than a recessive effect.

### Criteria

| Criteria | Complete | Incomplete |
|------------------------|------------------------|------------------------|
| **Data Inspection** | Correct sample/SNP counts and variable type identified. | Missing or incorrect counts or variable type. |
| **Allele Frequency Estimation** | Correct allele and minor allele frequencies computed. | Frequencies missing or wrong. |
| **Hardy–Weinberg Equilibrium Test** | Correct PLINK command and p-value extraction in R. | PLINK command or extraction incorrect/missing. |
| **Genetic Association Test** | Correct regressions, plots, coding, and interpretation. | Regression, plots, or interpretation missing/incomplete. |

### Submission Information

📌 Please review our [Assignment Submission Guide](https://github.com/UofT-DSI/onboarding/blob/main/onboarding_documents/submissions.md) for detailed instructions on how to format, branch, and submit your work. Following these guidelines is crucial for your submissions to be evaluated correctly.

#### Note:

If you like, you may collaborate with others in the cohort. If you choose to do so, please indicate with whom you have worked with in your pull request by tagging their GitHub username. Separate submissions are required.

------------------------------------------------------------------------

#### Submission Parameters

-   Submission Due Date: `11:59 PM – 16/03/2026`

-   Branch name for your repo should be: `assignment-1`

-   What to submit for this assignment:

    -   Populate this Quarto document (`assignment_1.qmd`).
    -   Render the document with Quarto: `quarto render assignment_1.qmd`.
    -   Submit both `assignment_1.qmd` and the rendered HTML file `assignment_1.html` in your pull request.

-   What the pull request link should look like for this assignment: `https://github.com/<your_github_username>/gen_data/pull/<pr_id>`

    -   Open a private window in your browser. Copy and paste the link to your pull request into the address bar. Make sure you can see your pull request properly. This helps the technical facilitator and learning support team review your submission easily.

------------------------------------------------------------------------

Checklist:

-   Created a branch with the correct naming convention.
-   Ensured that the repository is public.
-   Reviewed the PR description guidelines and adhered to them.
-   Verified that the link is accessible in a private browser window.
-   Confirmed that both `assignment_1.qmd` and `assignment_1.html` are included in the pull request.

If you encounter any difficulties or have questions, please don't hesitate to reach out to our team via our Slack help channel. Our technical facilitators and learning support team are here to help you navigate any challenges.
